Rank Product Tests

### 2021-07-27 R/rank_product_md.Rmd
fixed coding error

## Finals Data
# you'll probably have to change these paths to get this to work
load("rp_data.rdata")
```

# About the Rank Product 
The Rank-Product test is a nonparametric test developed with the use in replicated gene expression experiments in mind. The idea is that we have $k$ sets of rankings for $n$ genes and want to know if a gene is differentially expressed. Outside of biology, this test also has applications in statistical meta-analysis and feature selection.

The rank-product $RP$ statistic is simply computed as the geometric mean of the rankings $r$ across the $k$ replicates for item $i = 1, 2, \dots, n$:

@@

@@

## Function to Generate Rank Permutations

```{r}
assign_ranks <- function(c = 8) {
  # randomly assign ranks in 3 events & compute total
  # ranking in an event is considered to be independent from ranking in another event
  output <- data.frame(climber = sample(LETTERS, c, replace = FALSE)
                       , speed = sample(1:c, c, replace = FALSE)
                       , bould = sample(1:c, c, replace = FALSE)
                       , lead = sample(1:c, c, replace = FALSE)
                       ) %>%
    mutate(total = speed*bould*lead) %>%
    arrange(total)

@@

# Computing the Rank-Product For Finalists
We start with the lower $n$ case

In the Combined-Event World Cup Climbing competitions, there are $n = 6$ competitors. We compute the geometric mean of their rankings for the three events:

## Men

```{r, echo = FALSE}
knitr::kable(m_fin_rp)
```

## Women

```{r, echo = FALSE}
knitr::kable(f_fin_rp)
```

```{r}
set.seed(80085)
n_perm <- 1000
# Rank Product for Finals (n = 6, k = 3)
fin_perms <- lapply(1:n_perm, function(x) assign_ranks(6) %>% mutate(rp = total^(1/3))
                    )

# list of rank-products for each total rank
rp_sets <- lapply(1:6, function(r) sapply(fin_perms, function(x) x$rp[r])
                  )
```

## Men's Finals
```{r}
E_m <- sapply(1:6, function(x) mean(m_fin_rp$rp[x] >= rp_sets[[x]]) # false discovery rate: /x
              )
```

```{r, echo = FALSE}
data.frame(last = m_fin_rp$last
           , first = m_fin_rp$first
           , rank = m_fin_rp$rank
           , rp = m_fin_rp$rp
           , pval = E_m
           ) %>% 
  knitr::kable()
```

## Women's Finals
```{r}
E_f <- sapply(1:6, function(x) mean(f_fin_rp$rp[x] >= rp_sets[[x]]) # false discovery rate:/x
              )
```

```{r, echo = FALSE}
data.frame(last = f_fin_rp$last
           , first = f_fin_rp$first
           , rank = f_fin_rp$rank
           , rp = f_fin_rp$rp
           , pval = E_f
           ) %>% 
  knitr::kable()
```

# Computing the Rank-Product For Qualifiers

>>>>

## Finals Data
load("rp_data.rdata")
```

# About the Rank Product 
The Rank-Product test is a nonparametric test developed with the use in replicated gene expression experiments in mind. The idea is that we have $k$ sets of rankings for $n$ genes and want to know if a gene is differentially expressed. (A gene is declared differentially expressed if an observed difference or change in read counts or expression levels between two experimental conditions is statistically significant https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4827276/ ). Outside of biology, this test also has applications in statistical meta-analysis and feature selection.

If the rankings are balanced (that is, if each item has an equal number of replicates), then the rank-product $RP$ statistic for item $i$ can simply be computed as the geometric mean of the rankings $r_{ij}$ across the $k$ replicates for item $i = 1, \dots, n$ and replicate $j = 1,\dots, k$ :

@@

https://febs.onlinelibrary.wiley.com/doi/epdf/10.1016/j.febslet.2004.07.055

## Determining Significance Level

@@

## Function to Generate Rank-Product Permutations
```{r}
assign_ranks <- function(c = 8) {
  # randomly assign ranks in 3 events & compute product across events
  # ranking in an event is considered to be independent from ranking in another event
  output <- cbind(sample(1:c, c, replace = FALSE)
                  , sample(1:c, c, replace = FALSE)
                  , sample(1:c, c, replace = FALSE)
                  ) %>%
    apply(1, prod)

@@

```{r}
compare_rp <- function(rp_real, rp_perm) {
  # count how many of the permuted rank-products are smaller than the real rank-product
  sum(rp_real >= rp_perm) %>%
  return()
}

nsim <- 1000
nqual <- nrow(m_qual_rp)
rp_sims <- replicate(nsim, assign_ranks(nqual))
df <- data.frame(first = rep(NA, nqual)
                 , last = rep(NA, nqual)
                 , speed = rep(NA, nqual)
                 , bould = rep(NA, nqual)
                 , lead = rep(NA, nqual)
                 , total = rep(NA, nqual)
                 , count_above = rep(NA, nqual)
                 , E = rep(NA, nqual)
                 , q = rep(NA, nqual)
                 )

for (i in 1:nqual) {
  t <- m_qual_rp$total[i]                                            # experimental/real score
  #cat(m_qual_rp$first[i], m_qual_rp$last[i], " ", t, "\n")
  s <-  compare_rp(t, rp_sims) %>% sum()                             # count which simulated rp values <= real rp
  e <- s/nsim                                                        # avg expected value
  q <- e/i                                                           # q value (false discovery rate)
  #print(q)
  df[i,] <- c(m_qual_rp$first[i]
              , m_qual_rp$last[i]
              , m_qual_rp$speed[i]
              , m_qual_rp$bould[i]
              , m_qual_rp$lead[i]
              , m_qual_rp$total[i]
              , s
              , e
              , q
              )
}

df %>% knitr::kable()
```
